preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-19 Manuscript/manuscript.Rmd
Add discussion paragraph with new refs

**Background.** High-throughput chromosome conformation capture technology (Hi-C) revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, critical for cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. However, the high resolution of genomic annotations associated with boundaries, such as CTCF, members of cohesin complex, suggest they can inform the precise location of domain boundaries. Several methods attempted to leverage genome annotation data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary, and a much smaller number of boundaries than the rest of the genome (class imbalance).

@@

>>>>

**Background.** High-throughput chromosome conformation capture technology (Hi-C) revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, 3D domains critical for cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. However, the high resolution of genomic annotations associated with boundaries, such as CTCF, members of cohesin complex, suggest they can inform the precise location of domain boundaries. Several methods attempted to leverage genome annotation data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary, and a much smaller number of boundaries than the rest of the genome (class imbalance).

@@

*preciseTAD* requires both cell-type-specific chromatin conformation capture data (Hi-C in the current study) and genome annotation data. Hi-C data provides information about the approximate location of domain boundaries, while genome annotation data is used to train a model that can predict boundaries at the base-level resolution. The diversity of chromatin conformation capture technologies [@Kempfer:2019aa] and analysis methods [@Hansen:2019aa] limits the availability of harmonized cell-type-specific domain boundary information. Although the ENCODE, NIH Roadmap Epigenomics, FANTOM5, BLUEPRINT, and other members of the International Human Epigenome Consortium (IHEC) [@Stunnenberg:2016aa] have been actively cataloging cell-type-specific genome annotation datasets, these data remain very sparse. Efforts to systematically catalog chromatin conformation capture data [@Wang:2018ab] and impute missing cell-type-specific genome annotation information [@Schreiber:2020aa] started to appear. We envision the broader availability of matched cell-type-specific omics data will enable a more systematic analysis of domain boundaries using *preciseTAD*. Importantly, the *preciseTAD* model trained in one cell type can be used to predict domain boundaries in another cell line using only genome annotation information (CTCF, SMC3, RAD21, ZNF143) from that cell line, thus lowering the data availability requirements.
